---
title: "tidyhwsd: getting started"
author: "tidyhwsd authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidyhwsd: getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
library(tidyhwsd)
```

## Overview

`tidyhwsd` provides a tidyverse-friendly interface to the Harmonized World Soil Database (HWSD) v2.0.
It returns tibbles for point queries and `terra::SpatRaster` objects for spatial queries.

For a detailed tutorial, see `vignette("tutorial")`.

## Download the HWSD grid

The HWSD index grid (HWSD2.bil) is not shipped with the package. Download it once to a writable directory:

```{r}
# Replace with your preferred path
hwsd_download(ws_path = "~/data/HWSD2", verbose = TRUE)
```

You can set `WS_PATH` in `~/.Renviron` to make the path persistent between sessions:

```
WS_PATH=~/data/HWSD2
```

## Point queries (tibble output)

```{r}
pt <- hwsd_extract(
  coords = c(110, 40),
  param = c("SAND", "PH_WATER"),
  layer = "D1",
  ws_path = "~/data/HWSD2"
)
```

## Share-weighted synthesis

Use `hwsd_compose()` to synthesize values across components:

```{r}
props <- hwsd_props()
pt_syn <- hwsd_compose(
  coords = c(110, 40),
  param = c("SAND", "PH_WATER"),
  layer = "D1",
  ws_path = "~/data/HWSD2",
  props = props
)
```

## Bounding box queries (SpatRaster)

```{r}
sand_cn <- hwsd_extract(
  bbox = c(70, 18, 140, 54),
  param = "SAND",
  layer = "D1",
  ws_path = "~/data/HWSD2",
  tiles_deg = 5,
  cores = 4
)
```

## Further reading

- Full tutorial: `vignette("tutorial")`
- Units and aggregation: `vignette("units-and-aggregation")`
