---
title: "HWSD v2.0 units and aggregation"
author: "tidyhwsd authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HWSD v2.0 units and aggregation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
library(tidyhwsd)
```

## Data model (from the HWSD v2.0 manual)

HWSD v2.0 stores soil information per soil mapping unit (SMU). Each SMU can have
up to 12 components, with:

- **SEQUENCE**: order of components (sequence 1 is dominant)
- **SHARE**: percent share, summed to 100 within an SMU
- **LAYER**: D1 to D7 depth layers

`hwsd2_layers` is the component table; `hwsd_extract()` and `hwsd_compose()` map
SMU IDs to attributes using these fields.

## Units and codes

`hwsd_props()` provides a description and default aggregation rule for each
property, plus a `precision` field that encodes the rounding step used after
aggregation.

Common units (from the manual and field definitions):

- **Depth**: `TOPDEP`, `BOTDEP` in cm (D1 = 0-20 cm; ...; D7 = 150-200 cm)
- **Texture**: `SAND`, `SILT`, `CLAY` in percent by weight
- **Bulk density**: `BULK`, `REF_BULK` in g/cm3
- **Organic carbon**: `ORG_CARBON` in percent by weight
- **pH**: `PH_WATER` in water (-log(H+))
- **Total nitrogen**: `TOTAL_N` in g/kg
- **CEC / ECEC / TEB**: cmol/kg
- **Electrical conductivity**: `ELEC_COND` in dS/m
- **AWC**: `AWC` in mm/m (see manual table on USDA texture classes)

Code fields:

- `ROOT_DEPTH` is a **class code** (1-4; deep to very shallow)
- `ROOTS`, `IL`, `SWR` are **class codes** (see manual)
- `DRAINAGE` is a categorical class

## Aggregation behavior

`hwsd_extract()` returns **dominant component** values:

- SEQUENCE = 1 is preferred
- within SEQUENCE = 1, the largest SHARE is chosen
- if no SEQUENCE = 1 exists, the largest SHARE is chosen

`hwsd_compose()` performs **share-weighted synthesis** using rules from
`hwsd_props()`:

- numeric fields: `weighted_mean`
- categorical fields: `weighted_mode`
- ID-like fields: `drop`
- numeric outputs are rounded to the `precision` step

Example:

```{r}
props <- hwsd_props()
props$agg[props$property == "PH_WATER"] <- "dominant"

pt_dom <- hwsd_extract(
  coords = c(110, 40),
  param = c("PH_WATER", "SAND"),
  layer = "D1",
  ws_path = "D:/data/HWSD2"
)

pt_syn <- hwsd_compose(
  coords = c(110, 40),
  param = c("PH_WATER", "SAND"),
  layer = "D1",
  ws_path = "D:/data/HWSD2",
  props = props
)
```
